---
title: "Quantile-Quantile Plots"
author: "Jason Bryer"
date: 2021-03-01
draft: false
categories: ["R"]
tags: ["qqplots"]
#summary: This post explores how R generates random numbers and the use of seeds.
editor_options: 
  chunk_output_type: console
---

Quantile-quantile plots are a useful tool for determining whether a measure is normally distributed. From [Wikipedia](https://en.wikipedia.org/wiki/Q–Q_plot):

>> Q–Q (quantile-quantile) plot is a probability plot, which is a graphical method for comparing two probability distributions by plotting their quantiles against each other. First, the set of intervals for the quantiles is chosen. A point (x, y) on the plot corresponds to one of the quantiles of the second distribution (y-coordinate) plotted against the same quantile of the first distribution (x-coordinate). Thus the line is a parametric curve with the parameter which is the number of the interval for the quantile.

```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE, error=FALSE}
library(tidyverse)
library(reshape2)
```

Let's begin with a vector of values, `mpg` from the `mtcars` data frame. The plot below is a histogram of values, density of `mpg` in black, and normal distribution in blue.

```{r, warning = FALSE}
data(mtcars)
x <- mtcars$mpg
x <- x[order(x)] # Order the vector
ggplot(mtcars, aes(x = mpg)) + 
	geom_histogram(aes(y = ..density..), bins = 10, fill = 'grey50') +
	geom_density() +
	geom_function(fun = dnorm, 
				  args = list(mean = mean(x), sd = sd(x)),
				  color = 'blue') +
	xlim(mean(x) - 3 * sd(x), mean(x) + 3 * sd(x))
```

To calculate the theoretical quantiles, we start with a vector of length equal to the original data (i.e. n = `r length(x)` in this example). The `ppoints` generates a sequence of probability points that are equally distributed between 0 and 1.

```{r}
( points <- ppoints(length(x)) )
```

Using the `qnorm` function we can calucate the quantile for each of these values.

```{r}
( y <- qnorm(points) )
mean(y)
sd(y)
```

Note that the mean and standard deviation are 0 and 1 (within rounding error), respectively, as would be expected from a standard normal distribution. We will no convert our observed values to z-scores and see how they pair up with the theoretical quantiles.

```{r}
z <- (x - mean(x)) / sd(x)
df <- data.frame(ID = 1:length(x), Observed = x, Observed_z = z, Theoretical = y)
head(df)

ggplot(df, aes(x = Observed_z, y = 'Observed', xend = Theoretical, yend = 'Theoretical')) +
	geom_point(aes(x = Observed_z, y = 'Observed'), color = 'red') +
	geom_point(aes(x = Theoretical, y = 'Theoretical'), color = 'blue') +
	geom_segment() + ylab('') + xlab('')
```

The points in blue are the theoretical quantiles and each is matched to a point in red which is the observed z-score. If the data points were perfectly normal, then all the lines would be vertical. The following plot shows the distributions of these to vectors. 

```{r}
ggplot(data.frame(Observation = c(rep('Observed', length(z)),
								  rep('Theoretical', length(y))),
				  Value = c(z, y)), aes(x = Value, color = Observation)) +
	geom_density() +
	geom_rug() +
	scale_color_manual(values = c('Observed' = 'red', 'Theoretical' = 'blue')) +
	ylab('') + xlab('z-score')
```

We can see that the distribution does vary from the normal. The q-q plot provides another way of visualizing the departure from the normal distribution.

```{r}
ggplot(df, aes(x = Observed, y = Theoretical)) + 
	geom_point()
```

To add the line, we need to calculate the slope and intercept from two points that would fall exactly on the line if the distribution was perfectly normal.

```{r}
( x2 <- quantile(x, probs = c(0.25, 0.75)) ) # Using 25th and 75th percetile, but any two would work.
( y2 <- qnorm(c(0.25, 0.75)) )
slope <- diff(y2) / diff(x2)
int <- y2[1L] - slope * x2[1L]

ggplot(df, aes(x = Observed, y = Theoretical)) + 
	geom_point() +
	geom_abline(slope = slope, intercept = int)
```

